ANOVA Comparison Simulations: Skewed, Poisson Data

### 2016-12-07 simulations/simulation_skewed.Rmd
Add simulations for skewed discrete measures

---
title: "ANOVA Comparison Simulations: Gaussian Data"
author: "Kellie Ottoboni"
date: "`r Sys.Date()`"
output: pdf_document
---

@@

# Normal data

We assume the following linear data-generating process:

$$Y_{ij1} =\beta_0Y_{ij0} + \beta_{j} + \gamma_j Z_{ij} + \varepsilon_{ij}$$

for individuals $i = 1, \dots, n_j$, $j = 1, \dots, J$.
$\beta_0$ is the coefficient for the standard normally-distributed baseline measurement $Y_{i0}$, $\beta_j$ is the mean effect of being at site $j$, $Z_{ij}$ is the treatment level, $\gamma_j$ is the effect of treatment at site $j$, and $\varepsilon_{ij}$ is an error term.
We will assume that $\beta_0 = 1$.
The observed $(Y_{ij0}, \varepsilon_{ij})$ are independent across $i$ and $j$.

@@

```{r generate_sample}
generate_simulated_data <- function(gamma, effect, errors, n = c(16, 16, 16)){
  # Input:
  # gamma = the magnitude of the treatment effect
  # effect = "same effect" or "single site effect" - which sites have a tr effect > 0?
  # errors = "normal" or "heavy"
  # n = number of individuals at each site
  # Returns: a dataframe containing columns named Y1 (response), Y0 (baseline), Z (treatment), gamma_vec (treatment effect per individual), SITEID (stratum), site_effect (beta coefficient per individual), and epsilon (errors)

@@

  # Generate errors
  if(errors == "normal"){
    epsilon <- rnorm(N)
  } else {
    epsilon <- rt(N, df = 2)
  }

  # Generate covariates
  Y0 <- rnorm(N)
  Z <- rep(0:1, length.out = N)
  site_effect <- rep(beta, times = n)
  Y1 <- Y0 + gamma_vec*Z + site_effect + epsilon
  return(data.frame(Y1, Y0, Z, gamma_vec, SITEID, site_effect, epsilon))
}

@@

tmp <- generate_simulated_data(gamma = 0, effect = "same effect", errors = "normal")
design0_pvalues <- replicate(100, {
  tmp$epsilon <- rnorm(nrow(tmp))
  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$epsilon
  generate_simulated_pvalues(tmp)
})
design0_pvalues <- t(design0_pvalues)
colnames(design0_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
design0_power <- apply(design0_pvalues, 2, compute_power)
```

@@

## Design 1: Constant additive effect, normal errors

@@

tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "normal")
design1_pvalues <- replicate(100, {
  tmp$epsilon <- rnorm(nrow(tmp))
  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
  generate_simulated_pvalues(tmp)
})

@@

```{r design1_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design1_power, "Constant Additive Effect of 1, Normal Errors")

plot_pvalue_hist(design1_pvalues, "Constant Additive Effect of 1, Normal Errors")
```

## Design 2: Single site effect, normal errors

@@

tmp <- generate_simulated_data(gamma = 1, effect = "single site effect", errors = "normal")
design2_pvalues <- replicate(100, {
  tmp$epsilon <- rnorm(nrow(tmp))
  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
  generate_simulated_pvalues(tmp)
})
design2_pvalues <- t(design2_pvalues)
colnames(design2_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
design2_power <- apply(design2_pvalues, 2, compute_power)
```

```{r design2_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design2_power, "Effect of 1 at a Single Site, Normal Errors")

plot_pvalue_hist(design2_pvalues, "Effect of 1, Normal Errors")
```

## Design 3: Constant additive effect, heavy-tailed errors

Again, we find that the four power curves coincide while the curve for the stratified permutation test is below the others.
Here, the difference between the curves is not large.
Controlling for baseline $Y_0$ does not substantially help reduce variance.

tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "heavy")
design3_pvalues <- replicate(100, {
  tmp$epsilon <- rt(nrow(tmp), df = 2)
  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
  generate_simulated_pvalues(tmp)
})
design3_pvalues <- t(design3_pvalues)
colnames(design3_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
design3_power <- apply(design3_pvalues, 2, compute_power)
```

```{r design3_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design3_power, "Constant Additive Effect of 1, Heavy-tailed Errors")

plot_pvalue_hist(design3_pvalues, "Constant Additive Effect of 1, Heavy-tailed Errors")
```

## Design 4: single site effect, heavy-tailed errors

The noise from the heavy-tailed errors masks the treatment effect so much that controlling for baseline measures makes no difference.
There is almost no power to detect an effect using any of the five tests.

tmp <- generate_simulated_data(gamma = 1, effect = "single site effect", errors = "heavy")
design4_pvalues <- replicate(100, {
  tmp$epsilon <- rt(nrow(tmp), df = 2)
  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
  generate_simulated_pvalues(tmp)
})
design4_pvalues <- t(design4_pvalues)
colnames(design4_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
design4_power <- apply(design4_pvalues, 2, compute_power)
```

```{r design4_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design4_power, "Effect of 1 at a Single Site, Heavy-tailed Errors")

plot_pvalue_hist(design4_pvalues, "Effect of 1, Heavy-tailed Errors")
```

## Design 5: Unpredictive baseline

@@

@@

tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "normal")
design5_pvalues <- replicate(100, {
  tmp$epsilon <- rnorm(nrow(tmp))
  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
  tmp$Y1 <- 0.25*tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
  generate_simulated_pvalues(tmp)
})

design5_pvalues <- t(design5_pvalues)
colnames(design5_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
design5_power <- apply(design5_pvalues, 2, compute_power)
```

```{r design5_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design5_power, "Constant Additive Effect of 1, Normal Errors \n Unpredictive baseline")

plot_pvalue_hist(design5_pvalues, "Constant Additive Effect of 1, Normal Errors \n Unpredictive baseline")
```

# Imbalanced designs

## Design 6: Constant additive effect, normal errors

```{r design6, fig.height = 4}
set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC

tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "normal", n = c(8, 16, 24))
design6_pvalues <- replicate(100, {
  tmp$epsilon <- rnorm(nrow(tmp))
  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
  generate_simulated_pvalues(tmp)
})

design6_pvalues <- t(design6_pvalues)
colnames(design6_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
design6_power <- apply(design6_pvalues, 2, compute_power)
```

```{r design6_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design6_power, "Constant Additive Effect of 1, Normal Errors")

plot_pvalue_hist(design6_pvalues, "Constant Additive Effect of 1, Normal Errors")
```

## Design 7: Single site effect, normal errors

```{r design7, fig.height = 4}
set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC

tmp <- generate_simulated_data(gamma = 1, effect = "single site effect", errors = "normal", n = c(8, 16, 24))
design7_pvalues <- replicate(100, {
  tmp$epsilon <- rnorm(nrow(tmp))
  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
  generate_simulated_pvalues(tmp)
})
design7_pvalues <- t(design7_pvalues)
colnames(design7_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
design7_power <- apply(design7_pvalues, 2, compute_power)
```

```{r design7_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design7_power, "Effect of 1 at a Single Site, Normal Errors")

plot_pvalue_hist(design7_pvalues, "Effect of 1, Normal Errors")
```

# Heteroskedastic errors

One assumption of ANOVA is that errors are homoskedastic: they have the same variance across groups.
Suppose that this does not hold in our data.
Imagine we have three sites with 16 patients each, but they have different error distributions:
site 1 has standard normal errors,
site 2 has normally distributed errors with mean 0 and variance 2,
and site 3 has errors with a $t$-distribution on 2 degrees of freedom.

## Design 8: heteroskedastic errors, constant treatment effect

```{r design8, fig.height = 4}
set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
nn <- 16

tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "normal", n = rep(nn, 3))
design8_pvalues <- replicate(100, {
  tmp$epsilon[tmp$SITEID == 1] <- rnorm(nn)
  tmp$epsilon[tmp$SITEID == 2] <- rnorm(nn, sd = sqrt(2))
  tmp$epsilon[tmp$SITEID == 3] <- rt(nn, df = 2)
  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
  generate_simulated_pvalues(tmp)
})

design8_pvalues <- t(design8_pvalues)
colnames(design8_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
design8_power <- apply(design8_pvalues, 2, compute_power)
```

```{r design8_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design8_power, "Constant Additive Effect of 1, Heteroskedastic Errors")

plot_pvalue_hist(design8_pvalues, "Constant Additive Effect of 1, Heteroskedastic Errors")
```

## Design 9: heteroskedastic errors, single site effect

```{r design9, fig.height = 4}
set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
nn <- 16

tmp <- generate_simulated_data(gamma = 1, effect = "single site effect", errors = "normal", n = rep(nn, 3))
design9_pvalues <- replicate(100, {
  tmp$epsilon[tmp$SITEID == 1] <- rnorm(nn)
  tmp$epsilon[tmp$SITEID == 2] <- rnorm(nn, sd = sqrt(2))
  tmp$epsilon[tmp$SITEID == 3] <- rt(nn, df = 2)
  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
  generate_simulated_pvalues(tmp)
})
design9_pvalues <- t(design9_pvalues)
colnames(design9_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
design9_power <- apply(design9_pvalues, 2, compute_power)
```

```{r design9_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design9_power, "Effect of 1 at a Single Site, Heteroskedastic Errors")

plot_pvalue_hist(design9_pvalues, "Effect of 1, Heteroskedastic Errors")
```

>>>>

---
title: "ANOVA Comparison Simulations: Skewed, Poisson Data"
author: "Kellie Ottoboni"
date: "`r Sys.Date()`"
output: pdf_document
---

@@

# Poisson data

As in the other simulations that use Gaussian data, we assume the following linear data-generating process:

$$\tilde{Y}_{ij1} =\beta_0Y_{ij0} + \beta_{j} + \gamma_j Z_{ij} + \varepsilon_{ij}$$

for individuals $i = 1, \dots, n_j$, $j = 1, \dots, J$.
$\beta_0$ is the coefficient for the baseline measurement $Y_{i0}$, $\beta_j$ is the mean effect of being at site $j$, $Z_{ij}$ is the treatment level indicator,, $\gamma_j$ is the effect of treatment at site $j$, and $\varepsilon_{ij}$ takes the values $0$ or $0.5$ with equal probability.
We will assume that $\beta_0 = 1$.

Now, instead of $Y_{ij0}$ being normally distributed, we let these baseline measurements come from a truncated Poisson distribution with mean 4, with values greater than 10 censored and set to 10.
This is reflective of some survey data where individuals are asked to rate their experience on a scale from 1 to 10.
The data are not only discrete but also skewed:
more values are in the lower half of the range.

We don't observe $\tilde{Y}_{ij1}$ but $Y_{ij1}$, which is defined as

\begin{displaymath}
   Y_{ij1} = \left\{
     \begin{array}{ll}
       1 & \text{if } \tilde{Y}_{ij1} < 1\\
       10 & \text{if } \tilde{Y}_{ij1} > 10 \\
       \lfloor \tilde{Y}_{ij1} \rfloor & \text{otherwise}
     \end{array}
   \right.
\end{displaymath}

@@

```{r generate_sample}
generate_outcome <- function(baseline, tr_effect, site_effect, error){
  Y1 <- baseline + tr_effect + site_effect + error
  Y1 <- floor(Y1)
  Y1 <- pmax(pmin(Y1, 10), 1) 
  return(Y1)
}

generate_simulated_data <- function(gamma, effect, n = c(16, 16, 16)){
  # Input:
  # gamma = the magnitude of the treatment effect
  # effect = "same effect" or "single site effect" - which sites have a tr effect > 0?
  # n = number of individuals at each site
  # Returns: a dataframe containing columns named Y1 (response), Y0 (baseline), Z (treatment), gamma_vec (treatment effect per individual), SITEID (stratum), site_effect (beta coefficient per individual), and epsilon (errors)

@@

  # Generate covariates
  epsilon <- rbinom(N, 1, 0.5) - 0.5
  Y0 <- rpois(N, lambda = 4)
  Y0[Y0 > 10] <- 10
  Z <- rep(0:1, length.out = N)
  site_effect <- rep(beta, times = n)
  Y1 <- generate_outcome(Y0, gamma_vec*Z, site_effect, epsilon)
  return(data.frame(Y1, Y0, Z, gamma_vec, SITEID, site_effect, epsilon))
}

@@

tmp <- generate_simulated_data(gamma = 0, effect = "same effect")
design0_pvalues <- replicate(100, {
  tmp$epsilon <- rnorm(nrow(tmp))
  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
  tmp$Y1 <- generate_outcome(tmp$Y0, rep(0, nrow(tmp)), tmp$site_effect)
  generate_simulated_pvalues(tmp)
})
design0_pvalues <- t(design0_pvalues)
colnames(design0_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
design0_power <- apply(design0_pvalues, 2, compute_power)
```

@@

## Design 1: Constant additive effect

@@

tmp <- generate_simulated_data(gamma = 0.5, effect = "same effect")
design1_pvalues <- replicate(100, {
  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
  tmp$epsilon <- rbinom(nrow(tmp), 1, 0.5) - 0.5
  tmp$Y1 <- generate_outcome(tmp$Y0, tmp$Z*tmp$gamma_vec, tmp$site_effect, tmp$epsilon)
  generate_simulated_pvalues(tmp)
})

@@

```{r design1_plot, fig.height = 4, fig.align="center"}
p1 <- plot_power_curves(design1_power, "Constant Additive Effect of 0.5")
p1
plot_pvalue_hist(design1_pvalues, "Constant Additive Effect of 0.5")
```

## Design 2: Single site effect

@@

tmp <- generate_simulated_data(gamma = 0.5, effect = "single site effect")
design2_pvalues <- replicate(100, {
  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
  tmp$epsilon <- rbinom(nrow(tmp), 1, 0.5) - 0.5
  tmp$Y1 <- generate_outcome(tmp$Y0, tmp$Z*tmp$gamma_vec, tmp$site_effect, tmp$epsilon)
  generate_simulated_pvalues(tmp)
})
design2_pvalues <- t(design2_pvalues)
colnames(design2_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
design2_power <- apply(design2_pvalues, 2, compute_power)
```

```{r design2_plot, fig.height = 4, fig.align="center"}
p2 <- plot_power_curves(design2_power, "Effect of 0.5 in a Single Stratum")
p2
plot_pvalue_hist(design2_pvalues, "Effect of 0.5 in a Single Stratum")
```

## Design 3: Unpredictive baseline

@@

tmp <- generate_simulated_data(gamma = 0.5, effect = "same effect")
design3_pvalues <- replicate(100, {
  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
  tmp$epsilon <- rbinom(nrow(tmp), 1, 0.5) - 0.5
  tmp$Y1 <- generate_outcome(0.1*tmp$Y0, tmp$Z*tmp$gamma_vec, tmp$site_effect, tmp$epsilon)
  generate_simulated_pvalues(tmp)
})

design3_pvalues <- t(design3_pvalues)
colnames(design3_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
design3_power <- apply(design3_pvalues, 2, compute_power)
```

```{r design3_plot, fig.height = 4, fig.align="center"}
p3 <- plot_power_curves(design3_power, "Constant Additive Effect of 0.5 \n Unpredictive baseline")
p3
plot_pvalue_hist(design3_pvalues, "Constant Additive Effect of 0.5 \n Unpredictive baseline")
```

# Design 4: Constant additive effect, imbalanced design

tmp <- generate_simulated_data(gamma = 0.5, effect = "same effect", n = c(8, 16, 24))
design4_pvalues <- replicate(100, {
  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
  tmp$epsilon <- rbinom(nrow(tmp), 1, 0.5) - 0.5
  tmp$Y1 <- generate_outcome(tmp$Y0, tmp$Z*tmp$gamma_vec, tmp$site_effect, tmp$epsilon)
  generate_simulated_pvalues(tmp)
})

design4_pvalues <- t(design4_pvalues)
colnames(design4_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
design4_power <- apply(design4_pvalues, 2, compute_power)
```

```{r design4_plot, fig.height = 4, fig.align="center"}
p4 <- plot_power_curves(design4_power, "Constant Additive Effect of 0.5 \n Imbalanced Design")
p4
plot_pvalue_hist(design4_pvalues, "Constant Additive Effect of 0.5 \n Imbalanced Design")
```

# Design 5: Single site effect, imbalanced design

@@

tmp <- generate_simulated_data(gamma = 0.5, effect = "single site effect", n = c(8, 16, 24))
design5_pvalues <- replicate(100, {
  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
  tmp$epsilon <- rbinom(nrow(tmp), 1, 0.5) - 0.5
  tmp$Y1 <- generate_outcome(tmp$Y0, tmp$Z*tmp$gamma_vec, tmp$site_effect, tmp$epsilon)
  generate_simulated_pvalues(tmp)
})
design5_pvalues <- t(design5_pvalues)
colnames(design5_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
design5_power <- apply(design5_pvalues, 2, compute_power)
```

```{r design5_plot, fig.height = 4, fig.align="center"}
p5 <- plot_power_curves(design5_power, "Effect of 0.5 at a Single Site, Imbalanced Design")
p5 
plot_pvalue_hist(design5_pvalues, "Effect of 0.5 at a Single Site, Imbalanced Design")
```

```{r final_summary, fig.align = "center"}
grid.arrange(p1, p2, p4, p5)
```
